# Pmetrics Overview

## Software engines

There are three main software engines that Pmetrics controls.

* **IT2B** is the ITerative 2-stage Bayesian parametric population PK modeling
program. It is generally used to estimate parameter ranges to pass to
NPAG. It will estimate values for population model parameters under the
assumption that the underlying distributions of those values are normal
or transformed to normal.

* **NPAG** is the Non-parametric Adaptive Grid software. It will create a
non-parametric population model consisting of discrete support points,
each with a set of estimates for all parameters in the model plus an
associated probability (weight) of that set of estimates. There can be
at most one point for each subject in the study population. There is no
need for any assumption about the underlying distribution of model
parameter values.

* The **Simulator** is a semi-parametric Monte Carlo simulation software
program that can use the output of IT2B or NPAG to build randomly
generated response profiles (e.g. time-concentration curves) for a given
population model, parameter estimates, and data input. Simulation from a
non-parametric joint density model, i.e. NPAG output, is possible, with
each point serving as the mean of a multivariate normal distribution,
weighted according to the weight of the point. The covariance matrix of
the entire set of support points is divided equally among the points for
the purposes of simulation.

## Pmetrics control functions

<span class="r6">R6</span>

Pmetrics uses `PM_model` to create model objects and `PM_fit` to create
objects that combine the model with the data, ready to be run (fitted),
generating probability distributions for primary model parameters.

These functions replace the following Legacy functions: `ITrun`, `ERRrun`, 
`NPrun`.

`PMload` is aliased as `PM_load` for consistency with the above functions.
Similar to the Legacy version, it loads the results of a run into R to be 
accessible to the user for analysis, plotting, etc. Different from the Legacy
version, it loads the results into another R6 object, `PM_result` instead of
the current environment.  

Invoking the simulator in R6 is unchanged, i.e. use `SIMrun`.


<span class="legacy">Legacy</span>

Pmetrics has groups of R functions named logically to run each of these
programs and to extract the output. Again, these are extensively
documented within R by using the `help(command)` or `?command syntax`.

* `ITrun`, `ITparse`, `ERRrun`

* `NPrun`, `NPparse`

* `PMload`, `PMsave`, `PMreport`

* `SIMrun`, `SIMparse`

### Run functions

<span class="r6">R6</span>

Once a `PM_fit` object is created, which combines a model with a data file,
it can be run by using the syntax `$run()` to access the appropriate function
defined for the `PM_fit` object. 

```{r echo=T, eval=FALSE}
run1 <- PM_fit(model,data)
run1$run(options)
```


<span class="r6">R6</span> <span class="legacy">Legacy</span>

For IT2B and NPAG, the "run" functions generate batch files, which when
executed, launch the software programs to do the analysis. `$run(engine="err")`
or `ERRrun()` is a special implementation of IT2B designed to estimate 
the assay error polynomial coefficients from the data, when they cannot be calculated from assay validation data (using `makeErrorPoly()`) supplied by the
analytical laboratory. The batch files contain all the information
necessary to complete a run, tidy the output into a date/time stamped
directory with meaningful subdirectories, extract the information,
generate a report, and a saved Rdata file of parsed output which can be
quickly and easily loaded into R. On Mac (Unix) and Linux systems, 
the batch file automatically launches in a Terminal window. 
Prior to v1.9, on Windows systems, the batch file was launched manually, 
but as of v1.9, this manual step is no longer necessary. The execution of
the program to do the actual model parameter estimation is independent
of R, so that the user is free to use R for other purposes.

For the Simulator, the `SIMrun` function will execute the program
directly within R.

### Parse functions

<span class="r6">R6</span> <span class="legacy">Legacy</span>

For all programs, the "parse" functions will extract the primary output
from the program into meaningful R data objects. For IT2B and NPAG, this
is done automatically at the end of a successful run, and the objects
are saved in the output subdirectory as IT2Bout.Rdata or NPAGout.Rdata,
respectively. These functions generally run automatically and are not
necessary for the user to access.

### Saving and loading functions

<span class="r6">R6</span> 

After a successful IT2B or NPAG run, `PMload` or `PM_load` creates a 
`PM_result` object rather than loading run results into the current 
environment and suffixed with the run number.

```{r echo=T, eval=FALSE}
fit1 <- PM_load(1)
plot(fit1$op)
```

<span class="legacy">Legacy</span> 

For IT2B and NPAG, the `PM_load` function can be used to load either
of the above .Rdata files after a successful run. Objects will be loaded
into the current environment in R and suffixed with ".run", where "run" is
the run number.

```{r echo=T, eval=F}
PM_load(1)
plot(op.1)
```

<span class="update">Update</span>
`PMsave` is the
companion to `PMload` and can re-save modified objects to the .Rdata file.


### Report generation

<span class="r6">R6</span> <span class="legacy">Legacy</span>

The `PMreport` function is automatically run at the end of a successful
NPAG and IT2B run, and it will generate an HTML page with summaries of
the run, as well as the .Rdata files and other objects. The default
browser will be automatically launched for viewing of the HTML report
page. <!--It will also generate a .tex file suitable for processing by a
LATEX engine to generate a pdf report.--> See the [Pmetrics
Outputs](#outputs) section.

## Data manipulation functions

<span class="r6">R6</span> <span class="legacy">Legacy</span>

Within Pmetrics there are also functions to manipulate data .csv files
and process and plot extracted data.

* Manipulate data .csv files: `PMreadMatrix`, `PMcheck`, `PMwriteMatrix`,
`PMmatrixRelTime`, `PMwrk2csv`, `NM2PM`, `PMmb2csv`

* Process data: `makeAUC`, `makeCov`, `makeCycle`, `makeFinal`, `makeOP`, `makePTA`,
`makeErrorPoly`

* Plot data: `plot.PMcov`, `plot.PMcycle`, `plot.PMfinal`, `plot.PMmatrix`,
`plot.PMop`,`plot.PMsim`, `plot.PMvalid`, `plot.PMpta`

* Model selection and diagnostics: `PMcompare`, `plot.PMop` (with residual
option), `makeValid`, `plot.PMvalid`, `PMstep`

* Pmetrics function defaults: `setPMoptions`, `getPMoptions`

Again, all functions have extensive help files and examples which can be
examined in R by using the `help(command)` or `?command` syntax.