# How to use R and Pmetrics

In this section, we suggest a workflow to help you maintain organized
modeling projects.

## Setting up a Pmetrics project

When beginning a new modeling project, it is convenient to use the
command `PMtree`. This command will set up a new directory
in the current working directory named whatever you have included as the
"project name". 

```{r echo=T, eval=F}
PMtree("DrugX")
```

In the above example, a directory called "DrugX" will be created
in the current working directory in R, which you can check with the `getwd` function. Beneath the new DrugX directory, several subdirectories will
be also created.

* **Rscript** contains a skeleton R script to begin Pmetrics runs in the new project.
* **Runs** should contain all files required for a run (described next) and it will also contain the resulting numerically ordered run directories created after each Pmetrics NPAG or IT2B run.
* **Sim** can contain any files related to simulations
* **src** is a repository for original and manipulated source data files

You are free to edit this directory tree structure as you please, or make your
own entirely.

## Getting the required inputs to run Pmetrics

### R6
<span class="r6">R6</span>

There is a full tutorial encoded inside Pmetrics to teach new users the basic functionality of the whole package. To start that tutorial in R type:
```{r echo=T, eval=F}
library(Pmetrics)
PM_tutorial()
```
And follow the instructions prompted in the terminal.

To setup a R6 Pmetrics run, you must provide a `PM_data` and a `PM_model` objects.
It does not matter where the files (if there are any) that represent those files are located on your system.

Both objects provide a `$check` method to validate the consistency of each one:
```{r echo=T, eval=F}
my_model <- PM_model$new(...)
my_data <- PM_data$new(...)

# Scan both model and data for errors
my_model$check()
my_data$check()

```

The next step is to define a fitting problem, this is an object that represents a fitting proccess that will be solved by Pmetrics. On its minimum a fitting problem consists of a data object and a model object. This object is represented by the `PM_fit` class, let's see a couple of examples:
```{r echo=T, eval=F}
my_data <- PM_data$new(...)
my_simple_model <- PM_model$new(...)
my_complex_model <- PM_model$new(...)

#Here we're defining a fitting problem using the drug's data and a simple model:
simple_fit <- PM_fit$new(my_data, my_simple_model)

#Let's do the same but using the other model
complex_fit <- PM_fit$new(my_data, my_complex_model)
```

The fit object also contains a cumple of methods attached to it, i.e:

```{r echo=T, eval=F}
simple_fit$check() #This will check the consistency between the model and the data
complex_fit$save("complex_fit.rds") #This will save the fit to a file in your working directory

my_complex_fit_again <- PM_fit$load("complex_fit.rds") #This will load the fit again from disk
```

To see the full list of methods available see `?PM_fit`.


### Legacy
<span class="legacy">Legacy</span>

When you wish to execute a Pmetrics run, you must ensure that
appropriate Pmetrics model.txt and data.csv files are in the working
directory, i.e. the Runs subdirectory of the project directory. R can be
used to help prepare the data.csv file by importing and manipulating
spreadsheets (e.g. `read.csv`). The Pmetrics function `PMcheck` can be
used to check a .csv file or an R dataframe that is to be saved as a
Pmetrics data.csv file for errors. It can also check a model file for
errors in the context of a datafile, e.g. covariates that do not match.
`PMcheck(...,fix=T)` attempts to automatically rid data files of errors.
The function `PMwriteMatrix` can be used to write the R data object in
the correct format for use by IT2B, NPAG, or the Simulator.

You can also download sample data and scripts from the [Pmetrics
downloads](http://lapk.org/pmetrics.php) section of our
website. Edit prior versions of model files to make new model files.

## Using scripts to control Pmetrics
<span class="legacy">Legacy</span>

As you will see in the skeleton R script made by `PMtree` and placed in
the Rscript subdirectory, if this is a first-time run, the R commands to
run IT2B or NPAG are as follows. Recall that the "\#" character is a
comment character.

library(Pmetrics)

\#Run 1 - add your run description here

setwd("working directory")

NPrun() \#for NPAG or ITrun() for IT2B

The first line will load the Pmetrics library of functions. The second
line sets the working directory to the specified path. The third line
generates the batch file to run NPAG or IT2B and saves it to the working
directory.

**NOTE**: On Mac systems, the batch file will be automatically launched
in a Terminal window. On Windows systems prior to version 1.9, the batch
file must be launched manually by double clicking the *npscript.bat* or
*itscript.bat* file in the working directory. As of version 1.9, Windows
users no longer need to do this.

`ITrun` and `NPrun` are described in full detail via their help commands in R. At minimum, they require a data file and a model file. If the default names of "data.csv" and "model.txt" are used, they may be called with no arguments. Again, the data and model files must be in the current working directory, usually the Runs folder. 

Both functions return the full path of the output directory to
the clipboard. By default, runs are placed in folders numbered
sequentially, beginning with "1".

<span class="r6">R6</span>





## Loading results after a completed run
<span class="legacy">Legacy</span>

Now the output of IT2B or NPAG needs to be loaded into R, so the next
command does this.

`PMload(run_number)`

Details of these commands and what is loaded are described in the R
documentation (?PMload) and in the following section. The run\_number
should be included within the parentheses to be appended to the names of
loaded R objects, allowing for comparison between runs, e.g. `PMload(1)`.
Finally, at this point other Pmetrics commands can be added to the
script to process the data, such as the following.

<div class="script">
plot(final.1)\
plot(cycle.1)\
plot(op.1,type="pop") or plot(op.1\$pop1)\
plot(op.1) \#default is to plot posterior predictions for output 1\
plot(op.1,type="pop",resid=T)\
</div>

Of course, the full power of R can be used in scripts to analyze data,
but these simple statements serve as examples.

If you do not use the `PMtree` structure, we suggest that the R script
for a particular project be saved into a folder called "Rscript" or some
other meaningful name in the working directory. Folders are not be moved
by the batch file. Within the script, number runs sequentially and use
comments liberally to distinguish runs, as shown below.

library(Pmetrics)

\#Run 1 - Ka, Kel, V, all subjects

setwd("working directory")

NPrun() \#assumes model="model.txt" and data="data.csv"

PMload(1)

\...

Remember in R that the command `example(function)` will provide examples
for the specified function. Most Pmetrics functions have examples.

## Using Shiny to control Pmetrics

Shiny is an R package made by the same group who produces Rstudio. While
you are learning to use Pmetrics, you can use a graphical user interface
that teaches you how to build code. Currently we have implemented Shiny
interfaces for NPAG, IT2B and many plots in Pmetrics.

To use these Shiny interfaces, type `PMcode("x")`, where x is the function
you wish. Currently you have two choices: "run" and "plot".

`PMcode("run")` will launch the following dialog window in your default
browser. As you make your selections, you will see R code generated
which you can copy back to your script and execute. Simply close the
browser window when you are finished.

You can select from an NPAG or IT2B run, and then use the Data, Model,
and Run tabs to navigate to different options. With a valid Pmetrics
model and data file loaded, you will see a summary of the model in the
Current Model window.

If you use `PMcode("plot")` you will see the dialog window below. It will
allow you to select various Pmetrics objects that have been already
loaded with `PMload`, and offer you many plot options, as well as the
code to generate you selected plot.
