# Pmetrics Data Objects


After a successful IT2B or NPAG run, an R datafile is saved in the
output subdirectory of the newly created numerically ordered folder in
the working directory. After IT2B, this file is called "IT2Bout.Rdata",
and after NPAG it is called "NPAGout.Rdata". As mentioned in the
previous section, these data files can be loaded by ensuring that the
Runs folder is set as the working directory, and then using the Pmetrics
commands `PM_load()` in <span class="r6>R6</span> or `PMload()` for 
<span class="legacy">Legacy</span>.

There are several Pmetrics data objects contained within the Rdata files
which are loaded, making these objects available for
plotting and other analysis. 

<span class="r6>R6</span>

`PM_load()` returns a `PM_result` object. The fields in 

``` {r echo=T, eval=F}
res1 <- PM_load(1)

res1$

**Objects loaded by PMload(run\_num)**

```{r echo=F, results='asis'}
library(knitr)
tabObj <- read.csv("Data/Objects.csv",na.strings=".")
kable(tabObj)
```


Since R is an object oriented language, to access the observations in a
**PMop** object, for example, use the following syntax: op\$post1\$obs.

Note that you will place an integer corresponding to the run number
within the parentheses of the loading functions, e.g. PMload(1), which
will suffix all the above objects with that integer, e.g. op.1, final.1,
NPdata.1. This allows several models to be loaded into R simultaneously,
each with a unique suffix, and which can be compared with the
PMcompare() command (see [[Model Diagnostics]{.underline}](\l) below).

 